TacticToe: Learning the Game of Tactics

### 2018-01-26 tactictoe_jv.tex
prediction: wip

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 66.0\% of 
all theorems proven by a tactical script in the standard library.
This is to be compared with the 34.5\% proved in the same amount of time by 
Eprover in auto-schedule mode with 128 lemmas predicted and 
translated by HolyHammer.

@@

\subsection{The problem}\todo{}

Two ways of representing a formulas.

Defintion of a tactic in \holfour. In this research, we will take the definition
that a tactic takes a goal and returns a list of goals.

@@

The learning-based selection of relevant lemmas significantly improves the 
automation for hammers~\cite{BlanchetteGKKU16}. Therefore we propose to adapt 
one of the strongest hammer lemma selection 
methods: the modified distance-weighted \emph{$k$ nearest-neighbour} 
classifier~\cite{ckju-pxtp13,DudaniS76}.
In \tactictoe, we predict four different objects: theorems, terms, tactics and 
list of goals. We derive from tactic selection a prior policy for 
our search algorithm. Lists of goals (products of tactics) evaluate the success 
of the policy and adapt it accordingly. Term and theorems predictions are 
necessary to find the suitable arguments for abstracted tactics (see Section 
\ref{sec:synthesis}).

@@

%Mention explictly all the predictors that will be using this distance.
%tactics:
%theorems:
%terms:
%goal list: (product of tactics)
\begin{itemize}
\item tactic:\\
Given a goal $g$, the classifier selects a list of previously solved goals $l$
ordered by their similarity to $g$. Since each solved goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics. And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.
\item term:\\
Predict the closest term in a tactic that appears in the goal.

\item theorem:\\
Predict the theorem the closest to the current goal. Theorem dependencies are 
added during preselection.

@@

\subsection{Features for Terms and Goals}\label{sec:features}

We start by extracting the syntactic features that have been successfully used 
in premise selection from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single place holder $V$.
\item names of the variables
\end{itemize}

We found that the names of variables present in a term are particularly 
important for predicting tactics such as 
case splitting on a variable 
(\texttt{Cases\_on var}) or instantiation of a variable 
(\texttt{SPEC\_TAC var term}).

When a formula is represented as a goal, we also add a tag to each 
feature to remember whether it was extracted from the assumptions or the 
conclusion of a goal.
%We additionally extract the following features:

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llllll}
\toprule
 $c_{policy}$ & Width & Ortho & Eval & $c_{exploration}$ & Solved \\
\midrule
 2/3 & All & No & $E_{none}$ & 1 & 120\\
 1/2 & All & No & $E_{none}$ & 1 & 125\\
% 1/2 & All & Yes & $E_{none}$ & 1 & 130 \\
 1/3 & All & No & $E_{none}$ & 1 & 117\\
% 1/2 & Productive & No & $E_{none}$ & 1 & 120\\
% 1/2 & Productive & No & $E_{productive}$ & 1 & 127\\
\bottomrule
\end{tabular}
\caption{\label{tab:cost_param} on a local machine}
\end{table}

@@

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{llllll}
\toprule
  $c_{policy}$ & Width & Ortho & Eval & $c_{exploration}$ & Solved \\
\midrule
 1/2 & All & No & $E_{none}$ & 1 & 122\\
 1/2 & All & Yes & $E_{none}$ & 1 & 128\\
 1/2 & Productive & Yes & $E_{none}$ & 1 & 128\\
 1/2 & Productive & Yes & $E_{productive}$ & 1 & 134\\ 
 &  & &  & 4 & 130\\
 1/2 & Productive & Yes & $E_1$ & 1 & 127\\ 
 & & & & 4 & 126\\ 
 1/2 & Productive & Yes & $E_{10}$ & 1 & 136\\ 
 & & & & 4 & 137\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:learning} on the server}
\end{table}

@@

And we observe in Table~\ref{tab:featue_param} a 
slight increase compared to the default \tactictoe. We then included all 
vectors that contains tactics that can be rebuild from tokens of
tactics in our database. 
Our search algorithm proved two proofs less that if it had perfect prediction 
so it means that our prediction algorithm (policy) is doing a good  job at 
least in this early settings where proofs are easy. 
Suprisingly, being able to recombine tokens of tactics from the database is 
barely sufficient to prove more than half of the theorems in this dataset. The 
Therefore, additionally including in tactics theorems and terms that did not 
appear in previous tactics is essential. Creating from scratch other arguments 
may not help as much. This is why we restricted ourselves to predicting 
arguments of theorem list type and the term type during tactic synthesis. 
Finally our proof search algorithm sometimes fails even if it knows all the 
tactic necessary. The two main reasons are that either a crucial tactic exceeds 
its timeout or that the human proof is too deep to be rediscovered (more than 
10 steps long).

\subsubsection{Abstraction}
Here, we test the benefit of different kind of tactic abstraction. 
In all those experiments orthogonalization is performed with radius 20 but not 
orthogonalization with respect to metis.

All experiments in this table are using orthogonalization.
\begin{table}[t]
\centering\ra{1.3}
\small
\begin{tabular}{ll}
\toprule
  & Solved \\
\midrule
% holyhammer & 79 \\
 none & 122\\
 ortho & 129\\
 ortho + term & 126\\    
 ortho + theorem list & 160\\
 ortho + \metis & 162\\
 %ortho + metis + thmortho & 155\\
 %ortho + metis + stacortho & 160\\
 ortho + \metis + theorem list & 175\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:cfot_param}
Sucess rates of tactictoe with synthesis for different arguments.}
\end{table}
We want to know what the optimal number of predictions for tactics different 
from \metis since \metis has a special status in our algorithm is handled 
independently.

@@

\subsection{Slight improvements}
The final results could be improved by a few percent by including the 
following additional techniques that are already available.

\paragraph{Asynchronous calls to Eprover inside TacticToe}
The technique already discussed in Section~ would most likely improve the 
results by 2-3 percent.

\paragraph{Self-learning}
Until now, all our feature vectors were learned form human proofs. 
human proofs. We can also add to this tactics that were part of a proof found 
by our algorithm. To prevent duplication of effort, orthogonalization of those 
tactics is essential to have a beneficial effect (see Experiments).
Since recording and re-proving are intertwined, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own 
discovered proofs than from the human proof 
scripts~\cite{DBLP:conf/cade/Urban07}. It is also the first step toward 
self-reinforcement learning. Typically this technique improves the results by 
less than a percent.

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}     
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,original proofs}
\end{axis}
\end{tikzpicture}
\caption{Number of original and \tactictoe proofs with a number $x$ of tactic 
units. For comparing of the distribution, the number of \tactictoe proofs was 
scaled by the inverse of \tactictoe's success rate $\frac{1}{0.662}$.}
\end{figure}

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=20,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=oplen, y=solved] {data/tactictoe_by_oplen};
\addplot table[x=oplen, y=solved] {data/eprover_by_oplen};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

>>>>

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
TacticToe proves 66.2\% of 7164 theorems of the standard library where as 
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the 
result of both provers.

@@

\subsection{The problem}\todo{Maybe should }

Definitions of common concepts: goals, tactics, theorems, terms , etc

@@

The learning-based selection of relevant lemmas significantly improves the 
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe, we adapt 
one of the strongest lemma selection 
methods: the modified distance-weighted \emph{$k$ nearest-neighbour} 
classifier~\cite{ckju-pxtp13,DudaniS76} to predict four different objects: 
terms, goals, theorems and list of goals.

Before going into the details of the machine learning algorithm, we introduce
specificities associated with the prediction of each object: the data from 
which the feature vectors are build and the different use cases for the 
prediction.

We derive from tactic selection a prior policy for 
our search algorithm. Lists of goals (products of tactics) evaluate the success 
of the policy and adapt it accordingly. Term and theorems predictions are 
necessary to find the suitable arguments for abstracted tactics (see Section 
\ref{sec:synthesis}).

@@

%Mention explictly all the predictors that will be using this distance.
%tactics:
%theorems:
%terms:
%goal list: (product of tactics)
\begin{itemize}
\item term:\\
Predict the closest term in a tactic that appears in the goal.

\item :\\
Predict the theorem the closest to the current goal. Theorem dependencies are 
added during preselection.

\item goal:\\
Given a goal $g$, the classifier selects a list of previously solved goals $l$
ordered by their similarity to $g$. Since each solved goal is associated 
with a tactic in our database, the ordered list of goals induces a ordered  
list of associated tactics. And tactics are tried in this order during the 
proof search. Therefore, to order the tactics correctly, it is crucial to 
estimate the similarity between goals with mathematically relevant features.

@@

\subsection{Feature Extraction}\label{sec:features}
The four objects considered are different representation of mathematical 
formulas. Therefore, we start by describing which features we extract from the 
simplest representation terms. From there, we extend the extraction mechanism 
to goals theorems and lists of goals. Duplicated features are always removed 
so each object has an associated set of features.

Here are the different kind of feature we extract from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single place holder $V$.
\item names of the variables.
\end{itemize}

Since goals and theorems are represented by a couple of a list of terms 
(assumptions) and a term (conclusion). We compute their features by extracting 
the features of all terms describing them. Additionally, we distinguish between 
features of the assumptions and features of the conclusion by adding a 
different tag to elements of each set. For a list of goals, we simply construct 
the union of the features of each goal 
in the list.

%We found that the names of variables present in a term are particularly 
%important for predicting tactics such as 
%case splitting on a variable 
%(\texttt{Cases\_on var}) or instantiation of a variable 
%(\texttt{SPEC\_TAC var term}). 
%We additionally extract the following features:

@@

Width did include failing tactics in this experiments. $c_{exploration}$ = 1.
$E_{none}$.

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{ll}
\toprule
 $c_{policy}$ & Solved \\
\midrule
 $0.66$ & 120\\
 $0.5$ & 125\\
 $0.33$ & 117\\
% 1/2 & Productive & No & $E_{none}$ & 1 & 120\\
% 1/2 & Productive & No & $E_{productive}$ & 1 & 127\\
\bottomrule
\end{tabular}
\caption{\label{tab:cost_param} 
%on a local machine
}
\end{table}

@@

This experiments are performed with orthogonalization on and policy coefficient 
of $0.5$.

\begin{table}[ht]
\centering\ra{1.3}
\small
\begin{tabular}{lll}
\toprule
 Evaluation & $c_{exploration}$ & Solved \\
\midrule
% 1/2 & All & No & $E_{none}$ & 1 & 122\\
% 1/2 & All & Yes & $E_{none}$ & 1 & 128\\
 $E_{none}$ & 1 & 128\\
 $E_{productive}$ & 1 & 134\\ 
 & 4 & 130\\
 $E_1$ & 1 & 127\\ 
 & 4 & 126\\ 
 $E_{10}$ & 1 & 136\\ 
 & 4 & 137\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:learning} 
%on the server
}
\end{table}

@@

And we observe in Table~\ref{tab:featue_param} a 
slight increase compared to the default \tactictoe. We then included all 
vectors that contains tactics that can be rebuild from tokens of
tactics in our database. 
Our search algorithm proved two proofs less that if it had perfect prediction 
so it means that our prediction algorithm (policy) is doing a good  job at 
least in this early settings where proofs are easy. 
Suprisingly, being able to recombine tokens of tactics from the database is 
barely sufficient to prove more than half of the theorems in this dataset. The 
Therefore, additionally including in tactics theorems and terms that did not 
appear in previous tactics is essential. Creating from scratch other arguments 
may not help as much. This is why we restricted ourselves to predicting 
arguments of theorem list type and the term type during tactic synthesis. 
Finally our proof search algorithm sometimes fails even if it knows all the 
tactic necessary. The two main reasons are that either a crucial tactic exceeds 
its timeout or that the human proof is too deep to be rediscovered (more than 
10 steps long).In all those experiments orthogonalization is performed with 
radius 20.

\subsubsection{Abstraction}
Here, we test the benefit of different kind of tactic abstraction.
All experiments in this table are using orthogonalization.

\begin{table}[t]
\centering\ra{1.3}
\small
\begin{tabular}{ll}
\toprule
  & Solved \\
\midrule
% holyhammer & 79 \\
 none & 122\\
 ortho & 129\\
 ortho + term & 126\\    
 ortho + theorem list & 160\\
 ortho + \metis & 162\\
 %ortho + metis + thmortho & 155\\
 %ortho + metis + stacortho & 160\\
 ortho + \metis + theorem list & 175\\ 
\bottomrule
\end{tabular}
\caption{\label{tab:cfot_param}
Sucess rates of tactictoe with synthesis for different arguments.}
\end{table}
We want to know what the optimal number of predictions for tactics different 
from \metis since \metis has a special status in our algorithm is handled 
independently.

@@

\subsection{Slight improvements}\todo{maybe remove this subsection}
The final results could be improved by a few percent by including the 
following additional techniques that are already available.
The technique already discussed in Section~ would most likely improve the 
results by 2-3 percent. Reducing the exploration coefficient from 4 to 2 would 
also probably give a 3 percent increase in the number of proven theorems as 
shown by a 1 in 10 evaluation on the full library.

Until now, all our feature vectors were learned form human proofs. 
human proofs. We can also add to this tactics that were part of a proof found 
by our algorithm. To prevent duplication of effort, orthogonalization of those 
tactics is essential to have a beneficial effect (see Experiments).
Since recording and re-proving are intertwined, the 
additional data is available for the next proof search.
The hope is that the algorithm will improve faster by learning from its own 
discovered proofs than from the human proof 
scripts~\cite{DBLP:conf/cade/Urban07}. 
Self-learning improves the results by 
less than a percent.

%It is also the first step toward self-reinforcement learning.

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=3000,
  xmin=0, xmax=10,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\end{axis}
\end{tikzpicture}     
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=0.5*\textwidth,
  height=0.5*\textwidth,
  ymin=0, ymax=120,
  xmin=10, xmax=50,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=length, y=proofs] {data/tactictoe_proof_length};
\addplot table[x=length, y=proofs] {data/original_proof_length};
\legend{\tactictoe proofs,original proofs}
\end{axis}
\end{tikzpicture}
\caption{Number of original and \tactictoe proofs with a number $x$ of tactic 
units. For easier comparison of the distributions, the number of \tactictoe 
proofs was 
scaled by the inverse of \tactictoe's success rate $\frac{1}{0.662}$.}
\end{figure}

@@

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=20,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=oplen, y=solved] {data/tactictoe_by_oplen};
\addplot table[x=oplen, y=solved] {data/eprover_by_oplen};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original proof until length 20.}
\end{figure}
